Compositional Quantitative Resource Analysis of C Programs

### 2014-06-27 paper/main.tex
typos + remarks on logic soundness

$3.1|[y,z]| + 0.1|[0,y]|$
\\[.7\baselineskip] 
      {\bf t08}
    \end{center}
  \end{minipage}
%
\hfill
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  ${\color{blue}%
  n{<}0; \,  P(n,y) \vdash }$
  n=n+1;
  ${\color{blue}%
  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
  y=y+1000;
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  while (y>=100 && *){
    ${\color{blue}%
    y{>}99; \, 9 {+} P(n,y) \vdash }$
    y=y-100;
    ${\color{blue}%
    \dashv y{>}99; \, 14 {+} P(n,y) \vdash }$
    tick(5);
    ${\color{blue}%
    \dashv \cdot; \, 9 {+} P(n,y) }$
  }
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  tick(9);
  ${\color{blue}%
  \dashv \cdot; \, P(n,y) \vdash }$
}
   \end{lstlisting}
$59|[n,0]| {+} 0.05|[0,y]|$
\\[.7\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}

@@

In the example in~\pref{fig:xmplmax} we derive a precise runtime
bound on a program that searches a maximal element in an array.  The
cost metric that we use simply counts the assignments performed by the
program.  Hence, the resource consumption is closely related to the
number $A(0)$ of times the test \lstinline{a[i] > m} is true during the
execution.
%
We observe that $A(0)+1$ is the count of ``maximum candidates'' in the
array \lstinline{a} seen by the algorithm. We define
$$
 A(i) =  \# \{ k \mid i \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \} \, .
$$
where we write $\# S$ for the cardinal of the set $S$.  $A(0)$ is
obviously bounded by $N$, the size of the array.  So any automated tool
available would derive the linear bound $2 \cdot N$ for that program.
But with the expressivity of our  it is possible to use the
previous set cardinal directly and precisely tie the bound to the
initial contents of the array.

@@

The resource safety of a continuation $\cont$ is defined using three
assertions, one for each of the possible outcome of a program
statement.  We denote it by $\safeK n B R Q \cont$ and define
it to be
\begin{align*}
& \safe n B {\code{break}} \cont \\
\land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
\land& \safe n Q {\code{skip}} \cont.
\end{align*}
%
We can now define the \emph{semantic validity} of a judgement $\cdot; B; R
\vdash \htriple P S Q$ of the quantitative Hoare logic with
empty function context as follows
\begin{align*}
&\valid n B R P S Q := \\
&\qquad \forall m \, \cont \, x .\, (m \le n \land \safeK m {B+x} {R+x} {Q+x} \cont) \\
&\qquad \implies \safe m {P+x} S \cont.
\end{align*}
Note how the validity of a triple embeds the frame rule of
our logic, this refinement is necessary to have a stronger
induction hypothesis available during the proof.

Using this semantic validity  we can define the validity of
a function context $\Delta$ that we write $\validC n \Delta$ as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$.

XXXXX  Some more explanations.  Where is the difficulty? XXX

%
\begin{theorem}[Soundness of the logic]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable, then
  $$
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $$
\end{theorem}

XXXXX  Some more explanations.  Why n+1?XXX

%
\noindent
If $S$ is a complete program and $\Delta$ is empty by
expanding the definitions and remarking that $\validC n \Delta$
and $\safeK n B R Q \Kstop$ are true for every $n$ we get that 
$$
  \forall n .\, \safe n P S \Kstop.
$$
This means that for any starting state $\state$, $P(\state)$
provides enough resource for any run of the program $S$.  This
setting is actually the main use case of the previous theorem
which is stated as a stronger result to allow a proof by
induction.

@@

To find derivations in our Quantitative Logic automatically, the main
insight is to consider only assertions that match a certain shape.
The general form of a potential function that we consider is
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}|.
$$

@@

Since the different rules of~\pref{fig:auto} manage the
logical state in exactly the same way as the Quantitative
Logic, we will focus on the quantitative annotations in the
sequel.
%
As a convention, if $Q$ is a quantitative annotation we
will write $q_i$ for its $i$th coordinate.  The notation
$Q \pm n$ used in many rules defines a new context $Q'$
such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\,
q'_i = q_i$.  Finally, if a rule mentions $Q$ and $Q'$
and leaves the latter undefined at some index $i$ we
will implicitely assume that $q'_i = q_i$.

%
% Next paragraph is useless shit.
%

@@

From these three examples alone, one can see that our
automation follows very closely the rules of the logic.
The case of the update statement is more interesting
and specific to linear potential functions.
In the {\sc Q:IncP} rule the program updates a variable
$x$ to add it $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the
change of the program state.  We write $x'$ the value
of $x$ after the assignment ($x + y$). Since $x$ is
the only variable changed, only intervals of the form
$\inter u x$ and $\inter x u$ will be resized.
Note that for any $u$, $\inter x u$ will get smaller
with the update,  and if $x' \in \inter x u$ 
we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.
But $|\inter x {x'}| = |\inter 0 y|$, this means that
the potential $q'_{0y}$ in the postcondition can
be increased by $q_{xu}$ under the guard that
$x' \in \inter x u$.  Dually, the interval
$\inter v x$ can get bigger with the update.
And we know that $|\inter v {x'}| \ge y + |\inter v x|$,
so we decrease the potential of $\inter 0 y$
by $q_{vx}$.  Remark that the rule forces it only for
$v \not\in \mathcal U$, because if not, we know
that $x \le v$ so $|\inter v x| = 0$.

Another very technical rule is {\sc Q:Call}.  It needs to
account for the stack ``switching'' caused by the function
call, the argument/return value passing, and the preservation
of local variables.  We can sum up the requirements for
this rule as follows,
\begin{itemize}
\item
  any potential in the function pre/postcondition is equalized
  to its matching potential in the pre/precondition,
\item
  the potential of intervals $|\inter x y|$ is preserved
  if $x$ and $y$ are local variables,
\item
  and the unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are zeroed out in the postcondition.
\end{itemize}
If $x$ and $y$ are local variables and $\code{f}(x,y)$ is
called, the rule takes care to split the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
materialized in the equation $Q = P + S$.  Arguments in
the function precondition $(\Gamma_f, Q_f)$ are named
using a fixed vector $\Vargs$ of names different from
all program variables.  This prevents name conflicts
from happening and ensures that the substitution
$[\Vargs/\vec x]$ is meaningful.  Symmetrically,
we use the unique name $\Vret$ to represent the
return value in the function postcondition
$(\Gamma'_f, Q'_f)$.

>>>>

$3.1|[y,z]| + 0.1|[0,y]|$
\\[.7\baselineskip] 
      {\bf t08}
    \end{center}
  \end{minipage}
%
\hfill
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  ${\color{blue}%
  n{<}0; \,  P(n,y) \vdash }$
  n=n+1;
  ${\color{blue}%
  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
  y=y+1000;
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  while (y>=100 && *){
    ${\color{blue}%
    y{>}99; \, 9 {+} P(n,y) \vdash }$
    y=y-100;
    ${\color{blue}%
    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
    tick(5);
    ${\color{blue}%
    \dashv \cdot; \, 9 {+} P(n,y) }$
  }
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  tick(9);
  ${\color{blue}%
  \dashv \cdot; \, P(n,y) }$
}
   \end{lstlisting}
$59|[n,0]| {+} 0.05|[0,y]|$
\\[.7\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}

@@

In the example in~\pref{fig:xmplmax} we derive a precise runtime
bound on a program that searches a maximal element in an array.  The
cost metric that we use simply counts the assignments performed by the
program.  Hence, the resource consumption is closely related to the
number $A(1)$ of times the test \lstinline{a[i] > m} is true during the
execution.
%
We observe that $A(1)+1$ is the count of ``maximum candidates'' in the
array \lstinline{a} seen by the algorithm. We define
$$
 A(i) =  \# \{ k \mid i \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \} \, .
$$
where we write $\# S$ for the cardinal of the set $S$.  $A(1)$ is
obviously bounded by $N$, the size of the array.  So any automated tool
available would derive the linear bound $2 \cdot N$ for that program.
But with the expressivity of our logic it is possible to use the
previous set cardinal directly and precisely tie the bound to the
initial contents of the array.

@@

The resource safety of a continuation $\cont$ is defined using three
assertions, one for each of the possible outcome of a program
statement.  We denote it by $\safeK n B R Q \cont$ and define
it to be
\begin{align*}
& \safe n B {\code{break}} \cont \\
\land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
\land& \safe n Q {\code{skip}} \cont.
\end{align*}
%
We can now define the \emph{semantic validity} of a judgement $B; R
\vdash \htriple P S Q$ of the quantitative logic without function
context as follows
\begin{align*}
&\valid n B R P S Q := \\
&\qquad \forall m \, \cont \, x .\, (m \le n \land \safeK m {B+x} {R+x} {Q+x} \cont) \\
&\qquad \implies \safe m {P+x} S \cont.
\end{align*}
Note how the validity of a triple embeds the frame rule of
our logic, this refinement is necessary to have a stronger
induction hypothesis available during the proof.

Using the semantic validity of triples we can define the validity
of a function context $\Delta$, written $\validC n \Delta$, as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$.  A full
judgement that mentions a non-empty function context
$\Delta$ is in fact a guarded statement: it makes
assumptions on some functions' behavior.  The predicate
$\validC n \Delta$ gives the precise meaning of the
assumptions made.  It is also step-indexed to prove the
soundness of the {\sc L:Extend} rule by induction.
%
\begin{theorem}[Soundness of the logic]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable, then
  $$
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $$
\end{theorem}
%
\noindent
The difference $\delta = 1$ between the index in the triple
validity and the one in context validity is constrained by
the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
{\sc L:Call}, the language semantics makes one step and
proceeds with the function body, so we must have
$\delta \le 1$ to use the assumptions in $\Delta$.
For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
is a valid context for $n$ steps.  The induction hypothesis
in that case says that if $\Delta \cup \Delta'$ is valid
for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
So if we want to solve this goal by induction, it is
necessary that $\delta \ge 1$.  These two contraints force
$\delta$ to be exactly one in the theorem statement.

If $S$ is a complete program and $\Delta$ is empty, by
expanding the definitions we see that $\Delta$ is valid
for every $n$ and that $\Kstop$ is safe for every $n$, so
we get
$$
  \forall n .\, \safe n P S \Kstop.
$$
This means that from any starting state $\state$, $P(\state)$
provides enough resource for any run of the program $S$.  This
setting is actually the main use case of the previous theorem
which is stated as a stronger result to allow a proof by
induction.

@@

To find derivations in our quantitative logic automatically, the main
insight is to consider only assertions that match a certain shape.
The general form of a potential function that we consider is
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}|.
$$

@@

Since the different rules of~\pref{fig:auto} manage the
logical state in exactly the same way as the quantitative
logic, we will focus on the quantitative annotations in the
sequel.
%
As a convention, if $Q$ is a quantitative annotation we
will write $q_i$ for its $i$th coordinate.  The notation
$Q \pm n$ used in many rules defines a new context $Q'$
such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\,
q'_i = q_i$.  Finally, if a rule mentions $Q$ and $Q'$
and leaves the latter undefined at some index $i$ we
will implicitely assume that $q'_i = q_i$.

%
% Next paragraph is shit.
%

@@

From these three examples alone, one can see that our
automation follows very closely the rules of the logic.
The case of the update statement is more interesting
and specific to linear potential functions.
In the {\sc Q:IncP} rule the program updates a variable
$x$ to add it $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the
change of the program state.  We write $x'$ the value
of $x$ after the assignment ($x + y$). Since $x$ is
the only variable changed, only intervals of the form
$\inter u x$ and $\inter x u$ will be resized.
Note that for any $u$, $\inter x u$ will get smaller
with the update,  and if $x' \in \inter x u$ 
we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.
But $|\inter x {x'}| = |\inter 0 y|$, this means that
the potential $q'_{0y}$ in the postcondition can
be increased by $q_{xu}$ under the guard that
$x' \in \inter x u$.  Dually, the interval
$\inter v x$ can get bigger with the update.
And we know that $|\inter v {x'}| \le y + |\inter v x|$,
so we decrease the potential of $\inter 0 y$
by $q_{vx}$ to pay for this change.  Remark that
the rule forces it only for $v \not\in \mathcal U$,
because if not, we know that $x \le v$ so
$|\inter v x| = 0$.

Another very technical rule is {\sc Q:Call}.  It needs to
account for the stack ``switching'' caused by the function
call, the argument/return value passing, and the preservation
of local variables.  We can sum up the requirements for
this rule as follows,
\begin{itemize}
\item
  any potential in the function pre/postcondition is equalized
  to its matching potential in the call pre/postcondition,
\item
  the potential of intervals $|\inter x y|$ is preserved
  if $x$ and $y$ are local variables,
\item
  and the unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are zeroed out in the postcondition.
\end{itemize}
If $x$ and $y$ are local variables and $\code{f}(x,y)$ is
called, the rule takes care to split the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P + S$ and $Q' =
P' + S'$.  Arguments in the function precondition
$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts from happening and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function postcondition
$(\Gamma'_f, Q'_f)$.
